Search for WIMP Inelastic Scattering Off Xenon Nuclei With XENON100 Data

### 2017-01-19 Inelastic/analysis.tex
more_edits_captions_results_citations

As explained in Section~\ref{sec:intro}, the inelastic scattering of a WIMP with a $^{129}$Xe nucleus produces an energy deposit via a NR with subsequent emission of  
a 39.6\,keV de-excitation photon. The largest fraction of the energy released in the event is via an ER, due to the emitted photon which loses its energy in the LXe.
This represents an unusual signature compared to the one expected from an elastic scatter, and brings the signal region to overlap with the ER background region.
The selected region of interest (ROI) for this analysis surrounds the 39.6\,keV xenon line in the cS1-cS2 plane and is further divided into
sub-regions, as shown in Figures~\ref{fig:SR} and~\ref{fig:SR2}. {\textcolor{red} {Here more info on these regions, and the cS1 and cS2 boundaries?}

@@

\begin{figure}[t!]
  \includegraphics[width=\linewidth]{images/bkg_in_sr.png}
  \caption{Signal (1-9) and control (CR1 and CR2) regions for the inelastic WIMP-$^{129}$Xe interaction in the cS2 versus cS1 plane. {\textcolor{red} {I think we must explains the shape of these regions.}}}
  \label{fig:SR}
\end{figure}

@@

The distribution of an ER induced by the de-excitation photon in the cS1-cS2 plane  is simulated assuming a two dimensional normal probability distribution function (pdf), $f(cS1,cS2)$, 
described (apart from a constant normalisation factor) by the following equation:

\begin{multline}
	f(cS1,cS2)  = {\rm exp} \Big\{ -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \mu_{cS1})^2}{\sigma_{cs1}^2} + \\ 
	 \frac{(cS2 - \mu_{cS2})^2}{\sigma_{cs2}^2} - \frac{2\rho\cdot(cS1 - \mu_{cs1}) (cS2 - \mu_{cs2})} {\sigma_{cs1}\sigma_{cs2}} \Big] \Big\}
	%f(cS1,cS2) ~ = ~  \frac{1}{2 \pi \sigma_{cs1} \sigma_cs2 \sqrt{1-\rho^{2}} } exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \hat{cS1})^2}{\sigma_{cs1}^2} \Big] \Big)
\label{f:2dgaus}
\end{multline}
where $\mu_{cS1}$ and $\mu_{cS2}$ 
represent the average observed cS1 and cS2 signals given a 39.6\,keV ER, $\sigma_{cs1}$ and $\sigma_{cs2}$ are the standard deviation in cS1 and cS2 respectively,
while $\rho$ stands for the correlation between the cS1 and cS2 signals. The detector-related light yield $L_y$  at 39.6\,keV, necessary to evaluate the average number of prompt photons detected 
($\mu_{cS1}$), is obtained from the NEST model~\cite{NEST,Geant1,Geant2} fit to data collected with several $\gamma$-lines.  The average light yield at 39.6\,keV is 2.7\,PE/keV.

>>>>

As explained in Section~\ref{sec:intro}, the inelastic scattering of a WIMP with a $^{129}$Xe nucleus produces an energy deposit via a NR with subsequent emission of  
a 39.6\,keV de-excitation photon. The largest fraction of the energy released in the event is via an ER, due to the emitted photon which loses its energy in the LXe.
This represents an unusual signature compared to the one expected from an elastic scatter, and brings the signal region to overlap with the ER background region.
The selected region of interest (ROI) for this analysis surrounds the 39.6\,keV xenon line in the cS1-cS2 plane and is further divided into
sub-regions, as shown in Figures~\ref{fig:SR} and~\ref{fig:SR2}.  These sub-regions were defined such as to contain roughly a similar number of expected background events in each region. The control regions CR1 and CR2 are chosen to be as close as possible to the signal regions,  and are used as cross check of the background shape distribution.

@@

\begin{figure}[t!]
  \includegraphics[width=\linewidth]{images/bkg_in_sr.png}
  \caption{Signal (1-9) and control (CR1 and CR2) regions for the inelastic WIMP-$^{129}$Xe interaction in the cS2 versus cS1 plane. Their definition is given in the text.}
  \label{fig:SR}
\end{figure}

@@

The distribution of an ER induced by the de-excitation photon in the cS1-cS2 plane  is simulated assuming a two dimensional normal probability distribution function (pdf), $f(cS1_{er},cS2_{er})$, 
described (apart from a constant normalisation factor) by the following equation:

\begin{multline}
	f(cS1_{er},cS2_{er})  = {\rm exp} \Big\{ -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1_{er} - \mu_{cS1_{er}})^2}{\sigma_{cs1}^2} + \\ 
	 \frac{(cS2_{er} - \mu_{cS2_{er}})^2}{\sigma_{cs2}^2} - \frac{2\rho\cdot(cS1_{er} - \mu_{cs1}) (cS2_{er} - \mu_{cs2})} {\sigma_{cs1}\sigma_{cs2}} \Big] \Big\}
	%f(cS1,cS2) ~ = ~  \frac{1}{2 \pi \sigma_{cs1} \sigma_cs2 \sqrt{1-\rho^{2}} } exp \Big( -\frac{1}{2(1-\rho^2)} \Big[ \frac{(cS1 - \hat{cS1})^2}{\sigma_{cs1}^2} \Big] \Big)
\label{f:2dgaus}
\end{multline}
where $\mu_{cS1}$ and $\mu_{cS2}$ 
represent the average observed $cS1_{er}$ and $cS2_{er}$ signals given a 39.6\,keV ER, $\sigma_{cs1}$ and $\sigma_{cs2}$ are the standard deviation in $cS1_{er}$ and $cS2_{er}$ respectively,
while $\rho$ stands for the correlation between the cS1 and cS2 signals. The detector-related light yield $L_y$  at 39.6\,keV, necessary to evaluate the average number of prompt photons detected 
($\mu_{cS1}$), is obtained from the NEST model~\cite{NEST,Geant1,Geant2} fit to data collected with several $\gamma$-lines.  The average light yield at 39.6\,keV is 2.7\,PE/keV.

### 2017-01-19 Inelastic/results.tex
more_edits_captions_results_citations

\section{Results}
\label{sec:results}

Using an exposure of 34~kg of liquid xenon and 224.6~live days of data a yield of 764 events is observed  in the region of interest,
this is compatible with the expectation of $756 \, \pm \, 5^{(stat.)} \, \pm 55^{(syst.)}$ events from the background only hypothesis. 
Figure~\ref{fig:dataVSbkg} shows how these events are distributed in the region of interest, the bottom panel shows the ratio
between data and expected background, where the gray and orange shaded areas represent respectively statistical and systematic uncertainty 
on the background expectation.
%The distribution of events from data are compared to the background expectation in Figure~\ref{fig:dataVSbkg} and with the relative
%uncertainties.

This result is interpreted via a binned profiled likelihood approach by means of the test statistic $\tilde{q}$
and its asymptotic distributions described in \cite{asympt}. 
Assuming  an isothermal WIMP halo with a local density of $\rho_{\chi} \, = \, 0.3$~$GeV/cm^3$ , a local circular velocity of $v_0 \,= \, 220$~km/s, and a galactic escape velocity of $v_{esc} \, = \, 544$ km/s, 
\textcolor{blue}{other assumptions...,} a 90\% CL$_s$~\cite{cls} confidence level limit is  
computed on the spin dependent inelastic WIMP-nucleon cross section, $\sigma_{inel}$, as a function of the WIMP mass, $m_{\chi}$, and shown in Figure~\ref{fig:limits}.
The expected median sensitivity is reported with its relative one (green area) and two (yellow area) standard deviation uncertainty.
A  limit is set on $\sigma_{inel}$ to $3.3 \times 10^{-38} ~cm^{2}$ at 90\% CL$_s$ confidence level for a WIMP of mass 100~GeV. This limit is compared with
\textcolor{blue}{decide which other experiment to plot.}

\newpage
\begin{figure}[t!]
  \includegraphics[width=\linewidth]{images/data_vs_bkg.png}
  \caption{Results, comparison between data and expected background.}
  \label{fig:dataVSbkg}
\end{figure}

\begin{figure}[h]
  \includegraphics[width=\linewidth]{images/limit_reb.png}
  \caption{Observed and expected limits.}
  \label{fig:limits}
\end{figure}

\newpage

>>>>

\section{Results and Discussion}
\label{sec:results}

Using a fiducial volume containing 34\,kg of liquid xenon and 224.6~live days of data yields 764 events observed  in the region of interest.
This number is compatible with the expectation of $756 \, \pm \, 5 \,{\rm (stat)} \, \pm 55\, {\rm (syst)}$ events from the background only hypothesis. 
Figure~\ref{fig:dataVSbkg} shows the distribution of  events  in the region of interest, where the bottom panel displays the ratio
between data and expected background. The grey and orange shaded areas represent the statistical and systematic uncertainty 
on the background expectation, respectively. The expected signal for a WIMP mass of 100\,GeV/c$^2$, and normalised to a total of 50 events, is also shown.

%\newpage

\begin{figure}[t!]
  \includegraphics[width=\linewidth]{images/data_vs_bkg.png}
  \caption{Distribution of  observed events  in the region of interest (data points), along with the normalised distribution from calibration data (filled histogram). The bottom panel displays the ratio
between data and expected background, where the grey and orange shaded areas represent the statistical and systematic uncertainty 
on the background expectation, respectively. The expected signal for a WIMP mass of 100\,GeV/c$^2$ (blue dashed), and normalised to a total of 50 events, is also shown.}
  \label{fig:dataVSbkg}
\end{figure}

This result is interpreted via a binned profiled likelihood approach by means of the test statistic $\tilde{q}$
and its asymptotic distributions, as  described in \cite{asympt}. 
Assuming  an isothermal WIMP halo with a local density of $\rho_{\chi} \, = \, 0.3$\,GeV/cm$^3$, a local circular velocity of $v_0 \,= \, 220$\,km/s, and a galactic escape velocity of $v_{\rm{esc}} \, = \, 544$\,km/s, 
\textcolor{blue}{other assumptions...,} a 90\% CL$_s$~\cite{cls} confidence level upper limit on the spin-dependent inelastic WIMP-nucleon cross section as a function of WIMP mass is computed. We employ the nuclear structure factors as calculated in \cite{Baudis:2013qla}, based on state-of-the-art large-scale shell-model calculations, with chiral effective field theory WIMP-nucleon currents.

\begin{figure}[h]
  \includegraphics[width=\linewidth]{images/limit_reb.png}
  \caption{Upper limit (black curve) on the spin-dependent, inelastic WIMP-nucleon cross section as a function of WIMP mass.  The expected median sensitivity (dashed curve) along with the relative one (green area) and two (yellow area) standard deviation uncertainty is also shown. This result is compared to the upper limit (at 90\% C.L.) obtained by the XMASS experiment (red curve)~\cite{Uchida:2014cnn}.}
  \label{fig:limits}
\end{figure}

Our result is shown in Figure~\ref{fig:limits}, together with the expected median sensitivity and its relative one and two  standard deviation uncertainty.
The most constraining upper limit is  $3.3 \times 10^{-38}$\,cm$^{2}$ (at 90\% CL$_s$ confidence level) is for a WIMP of mass 100\,GeV/c$^2$.

This results is compared to the one obtained by the XMASS experiment~\cite{Uchida:2014cnn}, a single phase liquid xenon detector, which used a fiducial volume containing 41\,kg of LXe and 165.9~live days of data. \textcolor{blue}{Decide which other experiment to also plot - DAMA LXe? However only XMASS uses the same form factors as we do.}

Our limit improves upon the XMASS result and constrains new parameter space for the spin-dependent, inelastic scattering cross section. While these upper limits are not competitive to spin-dependent, elastic scattering results, as obtained by XENON100~\cite{Aprile:2013doa} and LUX~\cite{Akerib:2016lao} (with a cross section minimum of $\sim2 \times 10^{-40}$\,cm$^{2}$, at 90\% C.L.,  for a 100\,GeV/cm$^2$ WIMP), our results set the pathway for a sensitive search of inelastic WIMP-nucleus scattering in running or upcoming liquid xenon experiments such as XENON1T~\cite{Aprile:2015uzo}, XENONnT~\cite{Aprile:2015uzo},  LZ~\cite{Akerib:2015cja}, and DARWIN~\cite{Aalbers:2016jon}. In these larger detectors, with lower intrinsic backgrounds from $^{85}$Kr and $^{222}$Rn decays, and improved self-shielding, the electronic recoil background will be reduced by a few orders of magnitude with respect to XENON100, and ultimately dominated by solar neutrino interactions~\cite{Baudis:2013qla}.

The discovery of this interaction channel would be a clear signature for a spin-dependent nature of the interaction, and would provide a potential handle  to constrain the WIMP mass~\cite{Baudis:2013bba}.

%\newpage
